---
title: "Differential Expression Vignette"
author: "Jack Humphrey"
date: "01/04/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


Here we show how to replicate our differential expression (DE) results using the data uploaded to Zenodo.


```{r zenodo, cache=TRUE, message=FALSE }

library(zen4R) #remotes::install_github("eblondel/zen4R")
library(tidyverse)
library(limma)
library(edgeR)
library(gt)

# find record
zenodo <- ZenodoManager$new(logger = "INFO")
rec <- zenodo$getRecordById("6385747")
files <- rec$listFiles(pretty = TRUE)

## download files
infolder <- "download_zenodo/"
dir.create(infolder, showWarnings = FALSE)
if(!all(files$filename %in% list.files(infolder))){
  rec$downloadFiles(path = infolder)
}

```


Each spinal cord region has a separate tar.gz archive. Today we will work with the Cervical spinal cord region.

We have to first unzip the archive:


```{r unzip, cache=TRUE}
# unzip the Cervical Spinal Cord

tissue <- "Cervical_Spinal_Cord"

system( paste0("cd ", infolder,"; tar -xzf ", paste0(tissue, ".tar.gz" ) ))
list.files(file.path(infolder,tissue) )

```


Now we can read in the data


```{r reading, cache=TRUE, message=FALSE}
 
counts_file <- file.path(infolder, tissue, paste0(tissue, "_gene_counts.tsv.gz") )
tpm_file <- file.path(infolder, tissue, paste0(tissue, "_gene_tpm.tsv.gz") )
metadata_file <- file.path(infolder, tissue, paste0(tissue, "_metadata.tsv.gz") )

gene_meta <- read_tsv("download_zenodo/gencode.v30.gene_meta.tsv.gz")

support_loc <- read_tsv(metadata_file) %>% as.data.frame()
tpm_loc <- read_tsv(tpm_file)
counts_loc <- read_tsv(counts_file)

row.names(support_loc) <- support_loc$rna_id

genes_loc <- counts_loc %>% select(-gene_name) %>% 
  column_to_rownames("ensembl_id")
tpm_loc <- tpm_loc %>% select(-gene_name) %>% 
  column_to_rownames("ensembl_id")
```


```{r}
support_loc %>% group_by(disease) %>% tally() %>% gt::gt()
```


We use a liberal cut-off for lowly expressed genes (median TPM  > 0) in order to capture genes originating from lowly abundant cell-types, such as motor neurons.


```{r filtering, cache=TRUE}
keep.exp <- rowSums(tpm_loc > 0) >= ceiling(0.5 * ncol(genes_loc)) 

genes_loc <- genes_loc[keep.exp, row.names(support_loc)]

print(nrow(genes_loc))

```


We perform variable selection for our linear model.


```{r modelling, cache=TRUE, message=FALSE, warning=FALSE}
support_loc$rin_squared <- support_loc$rin^2
support_loc$age_squared <- support_loc$age_rounded^2

cols <- c("sex", "site_id", "disease", "library_prep")
support_loc <- support_loc %>% mutate_at(cols, funs(factor(.)))
support_loc$disease <- factor(support_loc$disease, levels = c("Control", "ALS"))

support_loc$age_squared <- support_loc$age_rounded^2

design_formula <- as.formula("~ disease + sex + rin + rin_squared + age_rounded + age_squared + library_prep + site_id + gPC1 + gPC2 + gPC3 + gPC4 + gPC5 + pct_mrna_bases")

design_loc <- model.matrix(design_formula, data = support_loc)
  
stopifnot( nrow(support_loc) == nrow(design_loc))
```


We now perform differential expression using Limma Voom


```{r run-limma, cache=TRUE}
norm_loc <- calcNormFactors(genes_loc, method = "TMM") 

dge <- DGEList(counts=genes_loc, samples=support_loc, norm.factors = norm_loc)  

v <- voom(dge, design_loc)
vfit <- lmFit(v, design_loc)
efit <- eBayes(vfit)

res <- 
    topTable(efit, coef=2, number = Inf, adjust.method = "BH") %>% 
      as.data.frame() %>%
      rownames_to_column(var = "geneid") %>%
      left_join(gene_meta, by = "geneid") %>%
      janitor::clean_names() %>%
      as_tibble() %>%
  select(genename, geneid, everything() ) %>%
  arrange(adj_p_val)

table(res$adj_p_val < 0.05)
table(res$adj_p_val < 0.05 & abs(res$log_fc) > 1)
```


We have identified thousands of differentially expressed genes. What are the top associations by P-value?


```{r de-results, cache=TRUE}
head(res, 10) %>% gt::gt()
```


We can plot the results as a volcano plot.


```{r volcano, cache=TRUE}
highlight <- c("CCL18", "CHIT1", "GPNMB", "MNX1", "MOBP")

res %>%
  mutate(case = case_when(
  adj_p_val < 0.05 & log_fc >= 1 ~ "A",
  adj_p_val < 0.05 & log_fc <1 & log_fc >0 ~ "B",
  adj_p_val < 0.05 & log_fc < -1 ~ "C",
  adj_p_val < 0.05 & log_fc > -1 & log_fc < 0 ~ "D",
  TRUE ~ "E"
  )) %>%
  mutate(label = ifelse(genename %in% highlight, genename, "" )) %>%
  ggplot(aes(x = log_fc, y = -log10(p_value))) + 
  geom_point(aes(colour = case), size = 0.8) +
  ggrepel::geom_text_repel(aes(label = label), max.overlaps = Inf ) +
  scale_colour_manual(values = c( "red", "salmon", "navy", "dodgerblue3", "gray" )) +
  theme_bw() + 
  theme(axis.text = element_text(colour = "black")) +
  guides(colour = "none") +
  labs(y = expression(-log[10]~P), x = expression(log[2]~fold~change), title = "Cervical Spinal Cord", subtitle = "ALS vs Control" )
  

```

